Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2023-01-05 analysis/paper_draft3.Rmd
Update paper_draft3.Rmd

In turn, the inference of bias based in part on historical strike data is relevant to lawyers for deciding whether to argue a _Batson_ violation, proving a prima facie case, and ultimately persuading the trial judge that bias in the particular case is more likely than not (Figure \@ref(fig:batsonworkflowfig)). One challenge for incorporating historical data is possible incompatibility between current and historical strike data for a given attorney that may arise because of missing data or because attorney bias depends on certain trial-level characteristics (e.g., defendant race, charge severity).[may need fleshing out]

@@

```
In this paper, we extend a Bayesian approach to _Batson_ and similar challenges [@Kadane2021; @Kadane2018a; @Kadane2018b] to incorporate historical strike data in a manner that allows us to explicitly incorporate the degree of incompatibility between current and historical trials. In this approach, we use two steps. In step one, we specify a model of the peremptory-strike process in the court of interest that includes a bias parameter to which we assign an initial prior distribution. We use an attorney's strike data from past trials, as generated by that same peremptory-strike process, to estimate a posterior distribution for that bias parameter. We use the power prior [@Chen2000; @Ibrahim2015] to control how much the strike history from past cases affects the posterior distribution of the bias parameter. In step two, we estimate the posterior distribution of the bias parameter from the current case, using the posterior distribution developed with the historical information and the power prior.

@@

Statistical methods for _Batson_ and similar challenges aim to help measure how much prospective jurors' race, gender, or other cognizable-class membership affected an attorney's use of peremptory strikes against them. [editme] The goal of these methods is not to _prove_ bias, but rather to lend support to an inferrence of bias and a subsequent _Batson_ challenge.
The Bayesian approach here accounts for available historical strike data when estimating the posterior for the bias parameter. By incorporating the power prior, one can easily assess how sensitive that posterior estimate is to the available historical strike data. [editme] sentence on why this is appealing

Moreover, with a tool like our prototype app, attorneys can use the approach here in to better detect strike bias for _Batson_ and related challenges. The posterior distribution of the striking attorney's bias parameter can help the lawyer decide in real time whether to make a  _Batson_ challenge at all. If argued, the attorney can introduce that posterior distribution as relevant evidence for presenting a prima facie case and of ultimately proving illegal bias (Figure \@ref(fig:batsonworkflowfig)). To be sure, such evidence alone cannot account for the veracity of the striking attorney's proferred reasons. Besides, _Batson_ procedure allows any type of evidence, not just statistical evidence, for ultimately proving or disproving illegal bias in the particular case. Thus, as with any use of historical strike data, the approach here cannot always determine whether a _Batson_ challenge should succeed or fail.

>>>>

In turn, the inference of bias based in part on historical strike data is relevant to lawyers for deciding whether to argue a _Batson_ violation, proving a prima facie case, and ultimately persuading the trial judge that bias in the particular case is more likely than not (Figure \@ref(fig:batsonworkflowfig)). Ignoring historical strike data increases the risk of underpowered or inflated estimates of attorney strike bias, because of the low number of strikes per trial (typically 6, 10, or 15 strikes per party).

Relying on historical strike data, however, requires an approach to missingness in the historical data or to the possibility that attorney's strike bias depends on certain trial-level characteristics that vary across trials (e.g., defendant race, charge severity). As a result, there may be incompatibility between plausible values of any attorney's bias parameter based on strike data for an attorney in the current trial and based on historical strike data for that attorney. Simply pooling the the strike data from current and past trials, and analyzing such data together, ignores such incompatibility.

@@

In this paper, we extend a Bayesian approach to _Batson_ and similar challenges [@Kadane2021; @Kadane2018a; @Kadane2018b] to incorporate historical strike data in a manner that allows us to explicitly and transparently adjust for our assumptions about the degree of incompatibility between current and historical trials. In this approach, we use two steps. In step one, we specify a model of the peremptory-strike process in the court of interest that includes a bias parameter to which we assign an initial prior distribution. We use an attorney's strike data from past trials, as generated by that same peremptory-strike process, to estimate a posterior distribution for that bias parameter. We use the power prior [@Chen2000; @Ibrahim2015] to control how much the strike history from past cases affects the posterior distribution of the bias parameter. In step two, we estimate the posterior distribution of the bias parameter from the current case, using the posterior distribution developed with the historical information and the power prior.

@@

Statistical methods for _Batson_ and similar challenges aim to help measure how much prospective jurors' race, gender, or other cognizable-class membership affected an attorney's use of peremptory strikes against them. These statistical methods do not aim to  _prove_ bias alone, but rather to lend support to an inferrence of bias and a subsequent _Batson_ challenge.

The Bayesian approach here accounts for available historical strike data when estimating the posterior for the bias parameter. By incorporating the power prior, one can easily assess how sensitive that posterior estimate is to the available historical strike data. With a tool like our prototype app, attorneys can use the approach here in to better detect strike bias for _Batson_ and related challenges. The posterior distribution of the striking attorney's bias parameter can help the lawyer decide in real time whether to make a  _Batson_ challenge at all. If argued, the attorney can introduce that posterior distribution as relevant evidence for presenting a prima facie case and of ultimately proving illegal bias (Figure \@ref(fig:batsonworkflowfig)). To be sure, such evidence alone cannot account for the veracity of the striking attorney's proferred reasons. Besides, _Batson_ procedure allows any type of evidence, not just statistical evidence, for ultimately proving or disproving illegal bias in the particular case. Thus, as with any use of historical strike data, the approach here cannot always determine whether a _Batson_ challenge should succeed or fail.
